---
title: "Expected Test Set MSE, Bias/Variance Trade-Off"
author: "Evan L. Ray"
date: "September 23, 2019"
output:
  ioslides_presentation:
    smaller: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = TRUE)
library(dplyr)
library(ggplot2)
library(grid)
library(purrr)
library(gridExtra)
```

<style> 
  slides > slide > hgroup + article { 
    margin-top: 10px; 
  } 
</style>

## Running Example Set Up

Consider a polynomial regression problem where the data are generated from

$$
y_i = 0.001 + 0.005 x_i -0.005 x_i^2 + 0.0002 x_i^3 + \varepsilon_i \\ \varepsilon_i \sim \text{Normal}(0, 0.4^2)
$$

```{r, echo = FALSE, fig.height=4}
set.seed(12345)
beta_0 <- 1/1000
beta_1 <- 5/1000
beta_2 <- -5/1000
beta_3 <- 0.2/1000
sigma <- 4 * 100/1000
f_true <- function(x) {
  beta_0 + beta_1 * x + beta_2 * x^2 + beta_3 * x^3
}
x <- seq(from = -10, to = 30, length = 25)
y <- f_true(x) + rnorm(length(x), sd=sigma)

x_test <- 20
y_test <- f_true(x_test) + rnorm(length(x_test), sd=sigma)

functions_toplot <- data.frame(
  x = seq(from = -10, to = 30, length = 101)
) %>%
  mutate(
    f_val = f_true(x),
    func = "True f(x)"
  )

ggplot() +
  geom_point(
    data = data.frame(
      x = c(x),
      y = c(y),
      group = c(rep("Train",length(x)))),
    mapping = aes(x = x, y = y, fill = group, shape = group),
    size = 2,
    color = "white") +
  geom_line(data = functions_toplot, mapping = aes(x = x, y = f_val, color = func)) +
  scale_shape_manual("Points", breaks = c("Train"), values = c(22)) +
  scale_fill_manual("Points", breaks = c("Train"), values = c("cornflowerblue")) +
  scale_color_manual("Functions", breaks = c("True f(x)"), values = c("black")) +
  theme_bw()
```



## Polynomial Fits of Degree 1, 3, and 15

```{r, echo = FALSE, fig.height=4}
set.seed(12345)
y <- f_true(x) + rnorm(length(x), sd=sigma)
y_test <- f_true(x_test) + rnorm(length(x_test), sd=sigma)

set.seed(12345)

beta_0 <- 1/1000
beta_1 <- 5/1000
beta_2 <- -5/1000
beta_3 <- 0.2/1000
sigma <- 4 * 100/1000

f_true <- function(x) {
  beta_0 + beta_1 * x + beta_2 * x^2 + beta_3 * x^3
}

x <- seq(from = -10, to = 30, length = 25)
x_test <- 20
x_pred <- sort(c(x_test, seq(from = -10, to = 30, length = 101)))

get_stuff_to_plot <- function(i) {
  y <- f_true(x) + rnorm(length(x), sd=sigma)
  y_test <- f_true(x_test) + rnorm(length(x_test), sd=sigma)

  functions_toplot <- data.frame(
    x = x_pred
  ) %>%
    mutate(
      f_val = f_true(x),
      func = "True f(x)"
    )

  mfit1 <- lm(y ~ x)
  f_fitted1 <- function(x) {
    predict(mfit1, newdata = data.frame(x = x))
  }
  
  mfit3 <- lm(y ~ poly(x, 3))
  f_fitted3 <- function(x) {
    predict(mfit3, newdata = data.frame(x = x))
  }
  
  mfit20 <- lm(y ~ poly(x, 15))
  f_fitted20 <- function(x) {
    predict(mfit20, newdata = data.frame(x = x))
  }
  
  functions_toplot <-
    rbind(
      functions_toplot,
      functions_toplot %>%
        mutate(
          f_val = f_fitted20(x),
          func = "Degree 15 Fits"
        ),
      functions_toplot %>%
        mutate(
          f_val = f_fitted3(x),
          func = "Degree 3 Fits"
        ),
      functions_toplot %>%
        mutate(
          f_val = f_fitted1(x),
          func = "Degree 1 Fits"
        )
    ) %>%
    mutate(
      sample_num = i
    )
  
  return(functions_toplot)
}

fs_toplot<- purrr::map_dfr(1, get_stuff_to_plot)


ggplot() +
  geom_point(
    data = data.frame(
      x = c(x),
      y = c(y),
      group = c(rep("Train",length(x)))),
    mapping = aes(x = x, y = y, fill = group, shape = group),
    size = 2,
    color = "white") +
  geom_line(data = fs_toplot %>% filter(func != "True f(x)") %>% mutate(func = factor(as.character(func), levels = paste0("Degree ", c(1, 3, 15), " Fits"))),
    mapping = aes(x = x, y = f_val, group = paste0(func, sample_num), color=func))+#,
#    color = "cornflowerblue") +
  geom_line(data = fs_toplot %>% filter(func == "True f(x)") %>% select(x, f_val),
    mapping = aes(x = x, y = f_val)) +
#  stat_function(fun = f_true, color = "black") +
  scale_shape_manual("Points", breaks = c("Train", "Test"), values = c(21,22)) +
  scale_fill_manual("Points", breaks = c("Train", "Test"), values = c("black")) +
#  scale_color_viridis_d("Functions", begin = 0.2, end = 0.8, option = "A") +
  scale_color_manual("Functions", values = c("cornflowerblue", "purple", "red")) +
  theme_bw()# +
#  facet_wrap(~ func, ncol = 3)
```



## Polynomial Fits of Degree 1, 3, and 15

Here are fits to a second randomly generated training set.

```{r, echo = FALSE, fig.height=4}
set.seed(12345)
y <- f_true(x) + rnorm(length(x), sd=sigma)
y_test <- f_true(x_test) + rnorm(length(x_test), sd=sigma)
y <- f_true(x) + rnorm(length(x), sd=sigma)
y_test <- f_true(x_test) + rnorm(length(x_test), sd=sigma)

set.seed(12345)

beta_0 <- 1/1000
beta_1 <- 5/1000
beta_2 <- -5/1000
beta_3 <- 0.2/1000
sigma <- 4 * 100/1000

f_true <- function(x) {
  beta_0 + beta_1 * x + beta_2 * x^2 + beta_3 * x^3
}

x <- seq(from = -10, to = 30, length = 25)
x_test <- 20
x_pred <- sort(c(x_test, seq(from = -10, to = 30, length = 101)))

get_stuff_to_plot <- function(i) {
  y <- f_true(x) + rnorm(length(x), sd=sigma)
  y_test <- f_true(x_test) + rnorm(length(x_test), sd=sigma)

  functions_toplot <- data.frame(
    x = x_pred
  ) %>%
    mutate(
      f_val = f_true(x),
      func = "True f(x)"
    )

  mfit1 <- lm(y ~ x)
  f_fitted1 <- function(x) {
    predict(mfit1, newdata = data.frame(x = x))
  }
  
  mfit3 <- lm(y ~ poly(x, 3))
  f_fitted3 <- function(x) {
    predict(mfit3, newdata = data.frame(x = x))
  }
  
  mfit20 <- lm(y ~ poly(x, 15))
  f_fitted20 <- function(x) {
    predict(mfit20, newdata = data.frame(x = x))
  }
  
  functions_toplot <-
    rbind(
      functions_toplot,
      functions_toplot %>%
        mutate(
          f_val = f_fitted20(x),
          func = "Degree 15 Fits"
        ),
      functions_toplot %>%
        mutate(
          f_val = f_fitted3(x),
          func = "Degree 3 Fits"
        ),
      functions_toplot %>%
        mutate(
          f_val = f_fitted1(x),
          func = "Degree 1 Fits"
        )
    ) %>%
    mutate(
      sample_num = i
    )
  
  return(functions_toplot)
}

fs_toplot<- purrr::map_dfr(1:2, get_stuff_to_plot) %>%
  filter(
    sample_num == 2
  )


ggplot() +
  geom_point(
    data = data.frame(
      x = c(x),
      y = c(y),
      group = c(rep("Train",length(x)))),
    mapping = aes(x = x, y = y, fill = group, shape = group),
    size = 2,
    color = "white") +
  geom_line(data = fs_toplot %>% filter(func != "True f(x)") %>% mutate(func = factor(as.character(func), levels = paste0("Degree ", c(1, 3, 15), " Fits"))),
    mapping = aes(x = x, y = f_val, group = paste0(func, sample_num), color=func))+#,
#    color = "cornflowerblue") +
  geom_line(data = fs_toplot %>% filter(func == "True f(x)") %>% select(x, f_val),
    mapping = aes(x = x, y = f_val)) +
#  stat_function(fun = f_true, color = "black") +
  scale_shape_manual("Points", breaks = c("Train", "Test"), values = c(21,22)) +
  scale_fill_manual("Points", breaks = c("Train", "Test"), values = c("black")) +
#  scale_color_viridis_d("Functions", begin = 0.2, end = 0.8, option = "A") +
  scale_color_manual("Functions", values = c("cornflowerblue", "purple", "red")) +
  theme_bw()# +
#  facet_wrap(~ func, ncol = 3)
```



## Polynomial Fits of Degree 1, 3, and 15

Here's what fits look like across 100 randomly generated training samples.

```{r, echo = FALSE, fig.height=3}
beta_0 <- 1/1000
beta_1 <- 5/1000
beta_2 <- -5/1000
beta_3 <- 0.2/1000
sigma <- 4 * 100/1000

f_true <- function(x) {
  beta_0 + beta_1 * x + beta_2 * x^2 + beta_3 * x^3
}

x <- seq(from = -10, to = 30, length = 25)
x_test <- 20
x_pred <- sort(c(x_test, seq(from = -10, to = 30, length = 101)))

get_stuff_to_plot <- function(i) {
  y <- f_true(x) + rnorm(length(x), sd=sigma)
  y_test <- f_true(x_test) + rnorm(length(x_test), sd=sigma)

  functions_toplot <- data.frame(
    x = x_pred
  ) %>%
    mutate(
      f_val = f_true(x),
      func = "True f(x)"
    )

  mfit1 <- lm(y ~ x)
  f_fitted1 <- function(x) {
    predict(mfit1, newdata = data.frame(x = x))
  }
  
  mfit3 <- lm(y ~ poly(x, 3))
  f_fitted3 <- function(x) {
    predict(mfit3, newdata = data.frame(x = x))
  }
  
  mfit20 <- lm(y ~ poly(x, 15))
  f_fitted20 <- function(x) {
    predict(mfit20, newdata = data.frame(x = x))
  }
  
  functions_toplot <-
    rbind(
      functions_toplot,
      functions_toplot %>%
        mutate(
          f_val = f_fitted20(x),
          func = "Degree 15 Fits"
        ),
      functions_toplot %>%
        mutate(
          f_val = f_fitted3(x),
          func = "Degree 3 Fits"
        ),
      functions_toplot %>%
        mutate(
          f_val = f_fitted1(x),
          func = "Degree 1 Fits"
        )
    ) %>%
    mutate(
      sample_num = i
    )
  
  return(functions_toplot)
}

fs_toplot<- purrr::map_dfr(1:100, get_stuff_to_plot)


ggplot() +
  geom_line(data = fs_toplot %>% filter(func != "True f(x)") %>% mutate(func = factor(as.character(func), levels = paste0("Degree ", c(1, 3, 15), " Fits"))),
    mapping = aes(x = x, y = f_val, color = func, group = paste0(func, sample_num))) +
  geom_line(data = fs_toplot %>% filter(func == "True f(x)") %>% select(x, f_val),
    mapping = aes(x = x, y = f_val)) +
#  stat_function(fun = f_true, color = "black") +
  scale_shape_manual("Points", breaks = c("Train", "Test"), values = c(21,22)) +
  scale_fill_manual("Points", breaks = c("Train", "Test"), values = c("orange", "cornflowerblue")) +
#  scale_color_viridis_d("Functions", end = 0.8, option = "C") +
#  scale_color_manual("Functions") +#, breaks = c("Fitted f(x)", "True f(x)"), values = c("cornflowerblue", "black")) +
  scale_color_manual("Functions", values = c("cornflowerblue", "purple", "red")) +
  theme_bw() +
  facet_wrap(~ func, ncol = 3)
```

 * **Bias:** Average (across training sets) value of prediction minus true function value
     * For many values of $x$, Degree 1 fit is biased
     * Degree 3 and 15 fits are unbiased
 * **Variance:** Variability of predicted values across training sets
     * Degree 15 fit has high variance
     * Degree 1 and 3 fits have lower variance

## Performance at a test point

We focus on measuring performance of our models at a particular input value, say $x_0 = 20$.

```{r, echo = FALSE, fig.height=3}
set.seed(12345)
beta_0 <- 1/1000
beta_1 <- 5/1000
beta_2 <- -5/1000
beta_3 <- 0.2/1000
sigma <- 4 * 100/1000
f_true <- function(x) {
  beta_0 + beta_1 * x + beta_2 * x^2 + beta_3 * x^3
}
x <- seq(from = -10, to = 30, length = 25)
y <- f_true(x) + rnorm(length(x), sd=sigma)

x_test <- 20
y_test <- f_true(x_test) + rnorm(length(x_test), sd=sigma)

functions_toplot <- data.frame(
  x = seq(from = -10, to = 30, length = 101)
) %>%
  mutate(
    f_val = f_true(x),
    func = "True f(x)"
  )

ggplot() +
  geom_point(
    data = data.frame(
      x = c(x, x_test),
      y = c(y, y_test),
      group = c(rep("Train",length(x)), "Test")),
    mapping = aes(x = x, y = y, fill = group, shape = group),
    size = 2,
    color = "white") +
  geom_line(data = functions_toplot, mapping = aes(x = x, y = f_val, color = func)) +
#  stat_function(fun = f_true, color = "black") +
  scale_shape_manual("Points", breaks = c("Train", "Test"), values = c(21,22)) +
  scale_fill_manual("Points", breaks = c("Train", "Test"), values = c("orange", "cornflowerblue")) +
  scale_color_manual("Functions", breaks = c("True f(x)"), values = c("black")) +
  theme_bw()
```

## Performance for Degree 1

```{r, echo = FALSE, fig.height=2.5}
set.seed(12345)
beta_0 <- 1/1000
beta_1 <- 5/1000
beta_2 <- -5/1000
beta_3 <- 0.2/1000
sigma <- 4 * 100/1000
f_true <- function(x) {
  beta_0 + beta_1 * x + beta_2 * x^2 + beta_3 * x^3
}
x <- seq(from = -10, to = 30, length = 25)


y <- f_true(x) + rnorm(length(x), sd=sigma)

x_test <- 20
y_test <- f_true(x_test) + rnorm(length(x_test), sd=sigma)

functions_toplot <- data.frame(
  x = seq(from = -10, to = 30, length = 101)
) %>%
  mutate(
    f_val = f_true(x),
    func = "True f(x)"
  )

mfit <- lm(y ~ x)
f_fitted <- function(x) {
  predict(mfit, newdata = data.frame(x = x))
}

functions_toplot <-
  rbind(
    functions_toplot,
    functions_toplot %>%
      mutate(
        f_val = f_fitted(x),
        func = "Fitted f(x)"
      )
  )


ggplot() +
  geom_point(
    data = data.frame(
      x = c(x, x_test),
      y = c(y, y_test),
      group = c(rep("Train",length(x)), "Test")),
    mapping = aes(x = x, y = y, fill = group, shape = group),
    size = 2,
    color = "white") +
  geom_line(data = functions_toplot, mapping = aes(x = x, y = f_val, color = func)) +
#  stat_function(fun = f_true, color = "black") +
  scale_shape_manual("Points", breaks = c("Train", "Test"), values = c(21,22)) +
  scale_fill_manual("Points", breaks = c("Train", "Test"), values = c("orange", "cornflowerblue")) +
  scale_color_manual("Functions", breaks = c("Fitted f(x)", "True f(x)"), values = c("cornflowerblue", "black")) +
  theme_bw()
```

We record 3 things:

1. Difference between test observation and fitted value: $y_0 - \hat{y}_0$
    * Average of squared values across all train/test samples is **Expected test MSE**
2. Difference between fitted value and true function value: $\hat{y}_0 - f(x_0)$
    * Average across all train/test samples is the **Bias**
    * Variance across all train/test samples is the **Variance**
3. Difference between test observation and true function: $y_0 - f(x_0)$.
    * Variance across all test samples is the **Model Error** (same as $Var(\varepsilon)$)



## Performance on 10,000 samples

```{r}
get_nums <- function(i) {
  y <- f_true(x) + rnorm(length(x), sd=sigma)
  
  x_test <- 20
  y_test <- f_true(x_test) + rnorm(length(x_test), sd = sigma)
  
  mfit1 <- lm(y ~ x)
  f_fitted1 <- function(x) {
    predict(mfit1, newdata = data.frame(x = x))
  }
  y_0_minus_y_hat_0_1 <- (y_test - f_fitted1(x_test))
  y_hat_0_minus_f_x_0_1 <- f_fitted1(x_test) - f_true(x_test)
  y_0_minus_f_x_0_1 <- (y_test - f_true(x_test))

  mfit3 <- lm(y ~ poly(x, 3))
  f_fitted3 <- function(x) {
    predict(mfit3, newdata = data.frame(x = x))
  }
  y_0_minus_y_hat_0_3 <- (y_test - f_fitted3(x_test))
  y_hat_0_minus_f_x_0_3 <- f_fitted3(x_test) - f_true(x_test)
  y_0_minus_f_x_0_3 <- (y_test - f_true(x_test))

  mfit10 <- lm(y ~ poly(x, 15))
  f_fitted10 <- function(x) {
    predict(mfit10, newdata = data.frame(x = x))
  }
  y_0_minus_y_hat_0_10 <- (y_test - f_fitted10(x_test))
  y_hat_0_minus_f_x_0_10 <- f_fitted10(x_test) - f_true(x_test)
  y_0_minus_f_x_0_10 <- (y_test - f_true(x_test))
  
  return(
    data.frame(
      quantity = rep(c("y_0 - Fitted", "Fitted - f(x_0)", "Model Error"), times = 3),
      value = c(y_0_minus_y_hat_0_1, y_hat_0_minus_f_x_0_1, y_0_minus_f_x_0_1, y_0_minus_y_hat_0_3, y_hat_0_minus_f_x_0_3, y_0_minus_f_x_0_3, y_0_minus_y_hat_0_10, y_hat_0_minus_f_x_0_10, y_0_minus_f_x_0_10),
      degree = rep(c(1, 3, 15), each = 3)
    )
  )
}

res <- purrr::map_dfr(1:10000, get_nums)
```

```{r, fig.height = 4}
ggplot(
  data = res %>%
    mutate(
      quantity = factor(as.character(quantity), levels = c("y_0 - Fitted", "Fitted - f(x_0)", "Model Error"))
    ),
  mapping = aes(x = value)
) +
  geom_histogram(bins=20) +
  facet_grid(degree ~ quantity)
```

```{r, echo = FALSE}
res %>% group_by(degree) %>%
  summarize(
    Expected_test_MSE = mean(value[quantity == "y_0 - Fitted"]^2),
    Bias = mean(value[quantity == "Fitted - f(x_0)"]),
    Variance = var(value[quantity == "Fitted - f(x_0)"]),
    Model_Error = var(value[quantity == "Model Error"]),
    Bias2_Var_Model_Error = Bias^2 + Variance + Model_Error
  )
```


