---
title: "Example of matrix formulation of multiple regression"
output: pdf_document
classoption: landscape
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(width = 200)
```

One of my favorite data sets contains a bunch of information about movies, including how the movie scores on the Bechdel test.  A movie passes the Bechdel test if it satisfies 3 rules:

1. it has at least two women;
2. the women talk to each other; and
3. they talk to each other about something or someone other than a man.

The full data set contains the following variables:

 * `year` is the year the movie was released
 * `title` is the title of the movie
 * `bechdel_test` is a version of the results of the Bechdel test with 5 categories, according to users of the website www.bechdeltest.com: "nowomen" means there are not at least two women in the movie; "notalk" means there are at least two women in the movie, but they don’t talk to each other; "men" means there are at least two women in the movie, but they only talk to each other about men; "dubious" means there was some disagreement among users of bechdeltest.com about whether or not the movie passed the test; and "ok" means that the movie passes the Bechdel test.
 * `bechdel_test_binary` is a the results of the Bechdel test with 2 categories, according to the users of the website www.bechdeltest.com: "PASS" means that the movie passed the test (i.e., its value for `bechdel_test` is "ok"); "FAIL" means it did not pass the test (i.e., its value for `bechdel_test` is something other than "ok")
 * `budget` is the movie's approximate production budget, in the dollars of the year the movie was made.
 * `domgross` is the movie's domestic gross earnings (i.e., total earnings from the U.S.), in the dollars of the year the movie was made.
 * `intgross` is the movie's combined domestic and international gross earnings (i.e., total earnings both in the U.S. and internationally), in the dollars of the year the movie was made.
 * `budget_2013` is the same as `budget` but in inflation-adjusted 2013 dollars.
 * `domgross_2013` is the same as `domgross`, but in inflation-adjusted 2013 dollars.
 * `intgross_2013` is the same as `intgross`, but in inflation-adjusted 2013 dollars.
 * `imdb_rating` is the average rating for the movie by users of the website www.imdb.com, on a scale of 0 to 10 (higher ratings are better)
 * `num_imdb_ratings` is the number of distinct users of www.imdb.com who have rated the movie.
 * `mpaa_rating` is the MPAA rating for the movie, like PG or R.
 * `run_time_min` is the length of the movie in minutes.

\newpage

```{r, warning=FALSE, message=FALSE, echo = FALSE}
library(readr)
library(dplyr)

movies <- read_csv("http://www.evanlray.com/data/bechdel/bechdel.csv") %>%
  filter(mpaa_rating %in% c("G", "PG", "PG-13", "R")) %>%
  mutate(
    bechdel_test = factor(bechdel_test, levels = c("nowomen", "notalk", "men", "dubious", "ok"), ordered = FALSE),
    bechdel_test_binary = factor(bechdel_test_binary, levels = c("FAIL", "PASS"), ordered = FALSE),
    mpaa_rating = factor(mpaa_rating, levels = c("G", "PG", "PG-13", "R"), ordered = FALSE)
  )# %>%
#  select(
#    intgross_2013, 
#  )
```

Suppose we want to model a movie's international gross earnings in inflation-adjusted 2013 dollars (`intgross_2013`) based on the following 5 explanatory variables: `budget_2013`, `run_time_min`, `imdb_rating`, `mpaa_rating`, `bechdel_test_binary`

To start with, let's just see what the design matrix for a first model looks like:

```{r, warning=FALSE, message=FALSE, fig.height=7, fig.width=9}
movies_fit <- lm(intgross_2013 ~ budget_2013 + run_time_min + imdb_rating + mpaa_rating + bechdel_test_binary, data = movies)

movies %>% select(
  budget_2013, run_time_min, imdb_rating, mpaa_rating, bechdel_test_binary
) %>%
  head()
head(model.matrix(movies_fit))
```

In this design matrix the fifth column has the values $$x_{i4} = \begin{cases} 1 \text{ if movie $i$ is rated PG} \\ 0 \text{ otherwise} \end{cases}$$

