---
title: "Q-Q Plots"
output:
  pdf_document:
    keep_tex: true
geometry: margin=1.5cm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(gridExtra)
```

Setting: We have $n$ observations of a random variable, and we want to know if they follow a normal distribution.

```{r}
set.seed(9789426)
example_data <- data.frame(
  x = rnorm(100, 5, 2)
)
```

One approach, good enough for most purposes: a density plot or histogram.

```{r, fig.height=2.5}
ggplot(data = example_data, mapping = aes(x = x)) +
  geom_histogram(bins = 10, fill = "gray", mapping = aes(y = ..density..)) +
  geom_histogram(binwidth = 0.1, fill = "orange", mapping = aes(y = ..density..)) +
  geom_density(color = "cornflowerblue")
```

Another option is the Quantile-Quantile plot.  Here we compare our sample to a standard Normal(0, 1) distribution (the default):

```{r, fig.height=2.5}
ggplot(data = example_data, mapping = aes(sample = x)) +
  stat_qq() +
  stat_qq_line()
```

Each point in this plot is of the form

\begin{equation*}
(\text{$q$-th quantile of theoretical distribution}, \text{$q$-th quantile of sample})
\end{equation*}

For example, if we pick $q = 0.1$, we can find a point in the plot above as follows:

The $x$ axis coordinate is the value $x$ such that $P(Z \leq x) = 0.1$, where $Z \sim N(0, 1)$:

```{r}
qnorm(0.1, mean = 0, sd = 1)
```

The $y$ axis coordinate is the value $y$ such that 10 percent of our sample data are less than y:

```{r}
quantile(example_data$x, 0.1)
```

So, the point $(-1.281552, 2.570362)$ is in our Q-Q plot:

```{r, fig.height=2.5}
ggplot(data = example_data, mapping = aes(sample = x)) +
  stat_qq() +
  stat_qq_line() +
  geom_point(x = -1.281552, y = 2.570362, color = "orange", size = 3)
```

If the sample data come from a normal distribution, the points should fall close to a line: there is an approximately linear relationship between sample and theoretical quantiles.

If the data come from some other distribution, we will see some non-linear pattern in the plot.  Here are some examples based on samples from skewed distributions.

```{r, echo = FALSE, fig.height=3.5}
set.seed(9748239)
example_data <- data.frame(
  x_skewed_left = rbeta(100, 7, 1),
  x_skewed_right = rbeta(100, 1, 7)
)

p1 <- ggplot(data = example_data, mapping = aes(x_skewed_left)) +
  geom_density() +
  ggtitle("Density, Skewed Left")

p2 <- ggplot(data = example_data, mapping = aes(x_skewed_right)) +
  geom_density() +
  ggtitle("Density, Skewed Right")

p3 <- ggplot(data = example_data, mapping = aes(sample = x_skewed_left)) +
  stat_qq() +
  stat_qq_line() +
  ggtitle("Q-Q, Skewed Left")

p4 <- ggplot(data = example_data, mapping = aes(sample = x_skewed_right)) +
  stat_qq() +
  stat_qq_line() +
  ggtitle("Q-Q, Skewed Right")

grid.arrange(p1, p2, p3, p4, ncol = 2)
```




