---
title: "Example of fitted values and orthogonal projections"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(width = 200)
```

**This example will be due as part of your next homework assignment.  We're just getting a start on it in class together.**

### Model Statement

Suppose we use the model
\begin{align*}
y_i &= \beta + \varepsilon_i \\
\varepsilon_i &\sim \text{Normal}(0, \sigma^2)
\end{align*}

Also suppose we have $n = 2$ observations, and the observed response vector is $y = \begin{bmatrix} 1 \\ 2 \end{bmatrix}$.

#### (a) What is the design matrix $X$?

\vspace{1in}

#### (b) Find the hat matrix $H$.

\vspace{2in}

#### (c) Find the fitted values $\hat{y} = H y$.

\vspace{2in}

#### (d) Draw a figure showing $\mathcal{C}(X)$ (it is a line), $y$, and $\hat{y}$, clearly labelling each.  Connect $y$ and $\hat{y}$ with a line segment, and by drawing an appropriate right angle on your figure, illustrate that $\hat{y}$ is the orthogonal projection of $y$ onto $\mathcal{C}(X)$.

```{r, echo = FALSE, fig.width=4, fig.height=4, warning=FALSE}
library(ggplot2)
ggplot() +
  scale_x_continuous(limits = c(-3, 3)) +
  scale_y_continuous(limits = c(-3, 3)) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  xlab("Observation 1") +
  ylab("Observation 2") +
  theme_bw()
```