---
title: "Minimizing RSS"
---

## Simple Linear Regression with distinct x values -- Everything Is OK

Suppose we have 3 observations:

 * $y_1 = 2$, $x_1 = 1$ (first observation is in the first group)
 * $y_2 = 4$, $x_2 = 2$ (second observation is in the second group)
 * $y_3 = 5$, $x_3 = 3$ (third observation is in the second group)

Our model is

$$
y_i = \beta_0 + \beta_1 x_i + \varepsilon_i \\
\varepsilon_i \sim \text{Normal}(0, \sigma^2)
$$

The design matrix is $X = \begin{bmatrix} 1 & 1 \\ 1 & 2 \\ 1 & 3 \end{bmatrix}$

We know that we can use the following code to find the estimates of $\beta_0$ and $\beta_1$:

```{r}
X <- cbind(
  c(1, 1, 1),
  c(1, 2, 3)
)
y <- matrix(c(2, 4, 5))

beta_hat <- solve(t(X) %*% X) %*% t(X) %*% y
beta_hat
```

Here is a picture of the RSS as a function of $\beta_0$ and $\beta_1$, with our estimates $(\hat{\beta}_0, \hat{\beta}_1)$ shown with a red point:

$$RSS = \sum_{i = 1}^n (y_i - \hat{y}_i)^2 = \{2 - (\beta_0 + \beta_1 \cdot 1)\}^2 + \{4 - (\beta_0 + \beta_1 \cdot 2)\}^2 + \{5 - (\beta_0 + \beta_1 \cdot 3)\}^2$$


```{r, echo = FALSE, message = FALSE, warning=FALSE}
library(rgl)
library(mvtnorm)
library(plyr)
library(dplyr)
```

```{r, echo = FALSE, message = FALSE}
X1 <- unexpanded_X <- seq(from = -25, to = 25, length = 101)
X2 <- unexpanded_Y <- seq(from = -12.5, to = 12.5, length = 101)
plot_df <- as.data.frame(expand.grid(X = unexpanded_X, Y = unexpanded_Y))

plot_df$Z <- (2 - (plot_df$X + plot_df$Y * 1))^2 + 
  (4 - (plot_df$X + plot_df$Y * 2))^2 +
  (5 - (plot_df$X + plot_df$Y * 3))^2

plot_df$Z[plot_df$Z >= 200] <- 200

min_rss <- (2 - (beta_hat[1, 1] + beta_hat[2, 1] * 1))^2 + 
  (4 - (beta_hat[1, 1] + beta_hat[2, 1] * 2))^2 +
  (5 - (beta_hat[1, 1] + beta_hat[2, 1] * 3))^2

color_n <- 1000 # number of colors used

z_lim <- range(plot_df$Z)
z_range <- z_lim[2] - z_lim[1]
z_colorlut <- rev(rainbow(1000, start = 0, end = 4/6)) # height color lookup table
z_col <- z_colorlut[ floor(color_n * (plot_df$Z - z_lim[1])/z_range) + 1 ]

junk <- open3d()
z_max <- max(plot_df$Z)
#plot3d(X1, X2, xlim=c(0, 3), ylim=c(0, 3), zlim=c(0, z_max), zlab="obs. 3", xlab = "obs. 1", ylab = "obs. 2", mouseMode = "zAxis", type = "s")
plot3d(X1, X2, xlim=c(-25, 25), ylim=c(-12.5, 12.5), zlim=c(0, z_max), zlab="RSS", xlab = "B_0", ylab = "B_1", mouseMode = "zAxis")
surface3d(X1, X2, plot_df$Z, alpha = 0.5, col = z_col)
points3d(matrix(c(beta_hat[1, 1], beta_hat[2, 1], min_rss), nrow = 1), col = z_colorlut[1000], size = 8.0)

view3d(userMatrix = structure(c(0.476727247238159, -0.212129727005959, 0.853072345256805, 
0, 0.879015803337097, 0.106304869055748, -0.464790940284729, 
0, 0.00791021343320608, 0.971442461013794, 0.237143740057945, 
0, 0, 0, 0, 1), .Dim = c(4L, 4L)))


#plotids <- with(iris, plot3d(Sepal.Length, Sepal.Width, Petal.Length, 
#                  type="s", col=as.numeric(Species)))
rglwidget(elementId = "rss_ok", width = 1000, height = 700)
```




## Simple Linear Regression with one x value -- Everything Is Broken

Suppose we have 3 observations:

 * $y_1 = 2$, $x_1 = 2$ (first observation is in the first group)
 * $y_2 = 4$, $x_2 = 2$ (second observation is in the second group)
 * $y_3 = 5$, $x_3 = 2$ (third observation is in the second group)

Our model is

$$
y_i = \beta_0 + \beta_1 x_i + \varepsilon_i \\
\varepsilon_i \sim \text{Normal}(0, \sigma^2)
$$

The design matrix is $X = \begin{bmatrix} 1 & 2 \\ 1 & 2 \\ 1 & 2 \end{bmatrix}$

We know that there is not a unique $\hat{\beta}$ that minimizes RSS because the columns of $X$ are not linearly independent.

Here is a picture of the RSS as a function of $\beta_0$ and $\beta_1$:

$$RSS = \sum_{i = 1}^n (y_i - \hat{y}_i)^2 = \{2 - (\beta_0 + \beta_1 \cdot 2)\}^2 + \{4 - (\beta_0 + \beta_1 \cdot 2)\}^2 + \{5 - (\beta_0 + \beta_1 \cdot 2)\}^2$$

Note that there is no unique pair $(\beta_0, \beta_1)$ that minimizes RSS.

```{r, echo = FALSE, message = FALSE, warning=FALSE}
library(rgl)
library(mvtnorm)
library(plyr)
library(dplyr)
```

```{r, echo = FALSE, message = FALSE}
X1 <- unexpanded_X <- seq(from = -25, to = 25, length = 101)
X2 <- unexpanded_Y <- seq(from = -12.5, to = 12.5, length = 101)
plot_df <- as.data.frame(expand.grid(X = unexpanded_X, Y = unexpanded_Y))

plot_df$Z <- (2 - (plot_df$X + plot_df$Y * 2))^2 + 
  (4 - (plot_df$X + plot_df$Y * 2))^2 +
  (5 - (plot_df$X + plot_df$Y * 2))^2

plot_df$Z[plot_df$Z >= 200] <- 200

min_rss <- (2 - (beta_hat[1, 1] + beta_hat[2, 1] * 2))^2 + 
  (4 - (beta_hat[1, 1] + beta_hat[2, 1] * 2))^2 +
  (5 - (beta_hat[1, 1] + beta_hat[2, 1] * 2))^2

color_n <- 1000 # number of colors used

z_lim <- range(plot_df$Z)
z_range <- z_lim[2] - z_lim[1]
z_colorlut <- rev(rainbow(1000, start = 0, end = 4/6)) # height color lookup table
z_col <- z_colorlut[ floor(color_n * (plot_df$Z - z_lim[1])/z_range) + 1 ]

junk <- open3d()
z_max <- max(plot_df$Z)
#plot3d(X1, X2, xlim=c(0, 3), ylim=c(0, 3), zlim=c(0, z_max), zlab="obs. 3", xlab = "obs. 1", ylab = "obs. 2", mouseMode = "zAxis", type = "s")
plot3d(X1, X2, xlim=c(-25, 25), ylim=c(-12.5, 12.5), zlim=c(0, z_max), zlab="RSS", xlab = "B_0", ylab = "B_1", mouseMode = "zAxis")
surface3d(X1, X2, plot_df$Z, alpha = 0.5, col = z_col)
points3d(matrix(c(beta_hat[1, 1], beta_hat[2, 1], min_rss), nrow = 1), col = z_colorlut[1000], size = 8.0)
segments3d(rbind(matrix(c(-25, -0.5 * (-25 - 3.6667), min_rss), nrow = 1), matrix(c(25, -0.5 * (25 - 3.6667), min_rss), nrow = 1)), col = z_colorlut[1000], lwd = 8.0)


view3d(userMatrix = structure(c(0.476727247238159, -0.212129727005959, 0.853072345256805, 
0, 0.879015803337097, 0.106304869055748, -0.464790940284729, 
0, 0.00791021343320608, 0.971442461013794, 0.237143740057945, 
0, 0, 0, 0, 1), .Dim = c(4L, 4L)))


#plotids <- with(iris, plot3d(Sepal.Length, Sepal.Width, Petal.Length, 
#                  type="s", col=as.numeric(Species)))
rglwidget(elementId = "rss_broken", width = 1000, height = 700)
```



