---
title: "For Loops in R"
output:
  pdf_document:
    keep_tex: false
header-includes:
   - \usepackage{booktabs}
geometry: margin=1.5cm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

## Additional reading/source for this handout

https://r4ds.had.co.nz/iteration.html

## Goal: we want to repeat a task many times

## Example: Mean of each variable in a data frame

The `mtcars` data set comes with R and contains observations of 11 numeric variables for 32 cars.

```{r}
head(mtcars)
dim(mtcars)
ncol(mtcars)
```

Let's find the mean of each of those variables.


## Overall structure

```{r, eval = FALSE}
1. Allocate space where results will be stored, if necessary
2. for(object in vector_of_objects) {
    a. do some stuff based on object
    b. potentially, save some results in the space allocated in step 1
}
```

### In our example:

```
1. Allocate space to store the column means
2. For each column in the `mtcars` data frame,
    a. Find the mean for that column
    b. Store the column mean in the appropriate entry of the space allocated in step 1.
```

The above is probably a minimal level of detail for pseudo code; you could add in some more detail if you want.  Below are three ways of actually implementing the algorithm above in R.

\newpage

### Approach 1:

```{r}
var_means <- vector("numeric", ncol(mtcars)) # 1. allocate space for results
for(i in seq_len(ncol(mtcars))) { # 2. set up for loop
  print(i) # not an important part of our work; just showing what's happening
  var_means[[i]] <- mean(mtcars[[i]]) # a. and b. do something based on i, save results
}

var_means
```

### Approach 2: 

Basically the same thing, a few details changed:

```{r}
#1. allocate space to store results
var_means_df <- data.frame(
  var_name = rep(NA, ncol(mtcars)),
  var_mean = rep(NA, ncol(mtcars))
)

for(i in seq_along(mtcars)) { # 2. set up for loop
  print(i) # not an important part of our work; just showing what's happening
  var_means_df$var_name[i] <- colnames(mtcars)[i] # a. and b. do something based on i, save results
  var_means_df$var_mean[i] <- mean(mtcars[[i]]) # a. and b. do something based on i, save results
}

var_means_df
```

### Approach 3

More of the same, but iterating over strings instead of integers

```{r}
#1. allocate space to store results
var_means_df_2 <- data.frame(
  var_name = colnames(mtcars),
  var_mean = rep(NA, ncol(mtcars))
)

for(var_name in colnames(mtcars)) { # 2. set up for loop
  print(var_name) # not an important part of our work; just showing what's happening
  save_ind <- which(var_means_df_2$var_name == var_name) # a. and b. do something based on i, save results
  
  var_means_df_2[save_ind, "var_mean"] <- mean(mtcars[[var_name]]) # a. and b. do something based on i, save results
}

var_means_df_2
```
