---
title: "Model Comparison -- Example 1"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We have a data set with observations of four variables measuring advertising budgets and sales for a product in each of 200 markets (provided as part of the ISLR package; I think the data are made up):

 * `sales` is a measure of sales volume in thousands of units
 * `TV` is TV advertising budget
 * `radio` is radio advertising budget
 * `newspaper` is newspaper advertising budget

Below is R code for making plots displaying three separate simple linear regression fits to the data (the actual plots are on the other side of the page).  In all three plots/models, `sales` is the response variable; the explanatory variable is different for each model.

```{r, fig.height= 3}
library(readr) # for read_csv, which can read csv files from the internet
library(ggplot2) # for making plots
library(gridExtra) # for grid.arrange, which arranges the plots next to each other

Advertising <- read_csv("http://www.evanlray.com/data/islr/Advertising.csv")

p1 <- ggplot(data = Advertising, mapping = aes(x = TV, y = sales)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)

p2 <- ggplot(data = Advertising, mapping = aes(x = radio, y = sales)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)

p3 <- ggplot(data = Advertising, mapping = aes(x = newspaper, y = sales)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)

grid.arrange(p1, p2, p3, nrow = 1)
```

With your neighbors, discuss which of these models would you prefer to use for predicting sales and why.

Then answer the questions below:

#### Being as specific and concrete as possible, write down a rule for selecting your preferred model based only on *visual* characteristics of the plot.  (That is, your rule should not involve any calculations of numeric quantities).

\vspace{4cm}

#### Being as specific and concrete as possible, write down a rule for selecting your preferred model based only on a *quantitative* summary of the data.  You can describe how you would calculate your numeric summary of the data; if you'd like you can write down a formula.




 
 
 
 
 
 