---
title: "Model Comparison -- Example 2"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We have a data set with information about 38 cars.  Here we will look at two of the variables measured about these cars, their `Weight` (explanatory variable) and their fuel efficiency (response variable) as measured in miles per gallon (`MPG`, higher MPG is more fuel efficient).  The full data set is loaded in and plotted here.  In addition, I have highlighted in red a subset of these observations that I will use in fitting three candidate models below.

```{r, fig.height = 3, fig.width = 3, message=FALSE, warning=FALSE}
library(readr) # for read_csv, which can read csv files from the internet
library(ggplot2) # for making plots
library(gridExtra) # for grid.arrange, which arranges the plots next to each other
library(polynom) # for obtaining the third polynomial fit below

cars <- read_csv("http://www.evanlray.com/data/sdm4/Cars.csv")
cars_subset <- cars[c(1, 6, 8, 14, 15, 16, 21, 32, 33, 37), ] # 10 observations to use in getting fits below.

ggplot() + 
  geom_point(data = cars, mapping = aes(x = Weight, y = MPG)) +
  geom_point(data = cars_subset, mapping = aes(x = Weight, y = MPG), color = "red")
```

Below is R code for making plots displaying three separate polynomial regression fits to the 10 observations highlighted in red above: one with a degree 1 polynomial (i.e., a line), one with a degree 2 polynomial (a parabola), and one with a degree 9 polynomial.

```{r, fig.height = 3.75}
lm1 <- lm(MPG ~ Weight, data = cars_subset)
predict_1 <- function(x) {
  predict(lm1, data.frame(Weight = x))
}

p1 <- ggplot(data = cars_subset, mapping = aes(x = Weight, y = MPG)) + 
  geom_point(color = "red") +
  stat_function(fun = predict_1) +
  ggtitle("linear fit")

lm2 <- lm(MPG ~ Weight + I(Weight^2), data = cars_subset)
predict_2 <- function(x) {
  predict(lm2, data.frame(Weight = x))
}

p2 <- ggplot(data = cars_subset, mapping = aes(x = Weight, y = MPG)) + 
  geom_point(color = "red") +
  stat_function(fun = predict_2) +
  ggtitle("quadratic fit")

# Our degree 9 polynomial fit is not obtained from lm (although you could do that too)
# You don't need to know how to use the poly.calc function.
fit9 <- poly.calc(cars_subset$Weight, cars_subset$MPG)
print(fit9)
predict_9 <- as.function(fit9)

p3 <- ggplot(data = cars_subset, mapping = aes(x = Weight, y = MPG)) + 
  geom_point(color = "red") +
  stat_function(fun = predict_9, n = 1000001) +
  ylim(c(15, 40)) +
  ggtitle("degree 9 fit, zoomed in")

p4 <- ggplot(data = cars_subset, mapping = aes(x = Weight, y = MPG)) + 
  geom_point(color = "red") +
  stat_function(fun = predict_9, n = 100001) +
  ggtitle("degree 9 fit, zoomed out")

grid.arrange(p1, p2, p3, p4, nrow = 2, ncol = 2)
```

With your neighbors, discuss which of these models you would prefer to use for predicting MPG and why.

Then answer the questions below:

#### If you applied the same rules you developed in Example 1, which model would you choose?  Is that the most appropriate model?

\vspace{2cm}

#### Being as specific and concrete as possible, write down a rule for selecting your preferred model based only on *visual* characteristics of the plot.  (That is, your rule should not involve any calculations of numeric quantities).

\vspace{2cm}

#### Being as specific and concrete as possible, write down a rule for selecting your preferred model based only on a *quantitative* summary of the data.  You can describe how you would calculate your numeric summary of the data; if you'd like you can write down a formula.




 
 
 
 
 
 